\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel Space

### 2012-05-12 design.tex
more

In future work (Section~\ref{sec:future-new}), the arbiter may also automatically identify extra decision points, such as on conflicting shared memory accesses.

@@

At each decision point, Landslide creates a new node in the decision tree. It stores the TID of the thread that was chosen, the state of the scheduler (what threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.

@@

\begin{itemize}
	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process could have yet to reap a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well. For these cases an additional check is needed.
	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.\footnote{
		If the kernel is idling and the shell is not waiting for keyboard input, it is likely that the test case has deadlocked.}
	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
\end{itemize}

The states in the test lifecycle, and the rules for transitioning between them are as follows.

\begin{enumerate}
	\item {\bf Bootup.} Initial state. When the kernel is idling and the shell is waiting for keyboard input, advance state.
	\item {\bf Keyboard input.} Record the current thread count (the ``start population''). Generate keystrokes to cause the desired test case to run.
		When the current thread count is greater than the start population, advance state.
	\item {\bf Test running.} Begin constructing the decision tree here (since it is pointless to backtrack to before the test begins). When the current thread count equals the start population, advance state.
	\item {\bf Test cleanup.} The test processes have exited, but some housekeeping may yet remain. When the the kernel is idling and the shell is waiting for keyboard input, advance state.
	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack.
\end{enumerate}

@@

\begin{enumerate}
	\item {\bf Kernel panic bugs.} If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
	\item {\bf Use-after-free bugs.} Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range. If not, Landslide proclaims the access to be illegal.\footnote{In the same way, Landslide also detects calls to \texttt{free} on blocks that were already freed or never allocated at all.}
	\item {\bf Deadlock bugs.} If Landslide finds no runnable threads on the runqueue\footnote{Except for idle, if it exists.}, or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
\suspend{enumerate}

@@

\begin{itemize}
	\item {\bf Current thread not on runqueue.} If the currently-running thread is not stored on the runqueue, we identify it as runnable anyway (except in the special case of idle, as described below).
	\item {\bf Idle thread.} Some kernels may have an explicit idle thread and store it on the runqueue, with explicit code to skip over it if other threads are runnable. Hence, if the idle thread is on the runqueue (and/or the current thread), it is only runnable if no other threads are runnable.
	\item {\bf Yielding mutexes.} In kernels whose mutexes (or other synchronisation idioms) leave threads on the runqueue when they are notionally blocked, such threads are not runnable.
\end{itemize}

Using this notion of runnability, the happens-before relation is established as follows. Let $X$ be a transition and $Y$ be a subsequent transition in a branch that was just explored, and $T_Y$ be the thread associated with $Y$. We say that $X$ {\em enables} $Y$ if $T_Y$ was not runnable immediately before $X$, but was runnable immediately after. Then, a transition $A$ happens-before a transition $B$ if $A$ is the same thread as $B$, if $A$ enables $B$, or if $A$ happens-before some transition $C$ that enables $B$.

@@

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
	\subfloat[Race-free interleaving.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$C$ & & & \texttt{x=42;} \\
			$B$ & & \texttt{y=31337;} & \\
			\multirow{2}{*}{$A_1$} & \texttt{if (y==31337)} & & \\
			& \texttt{assert(x==42);} & & \\
			\cline{2-4}
		\end{tabular}
	} &
	\subfloat[Alternate interleaving explored with DPOR.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$C$ & & & \texttt{x=42;} \\
			$A_2$ & \texttt{if (y==31337)} & & \\
			$B$ & & \texttt{y=31337;} & \\
			\cline{2-4}
		\end{tabular}
	} \\
	\multicolumn{2}{c}{
	\subfloat[Buggy interleaving.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$B$ & & \texttt{y=31337;} & \\
			\multirow{2}{*}{$A_3$} & \texttt{if (y==31337)} & & \\
			& \texttt{assert(x==42);} & & \\
			$C$ & & & \texttt{x=42;} \\
			\cline{2-4}
		\end{tabular}
	}}
	\end{tabular}
	\caption{This sample code defeats a DPOR implementation that only uses the first evil ancestor if a decision point is not defined between the two lines of thread 1's code. The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next. In the second interleaving, $C$ is no longer a second evil ancestor of $A_2$, so the third interleaving is never explored, and the assertion failure is missed.}
	\label{fig:evil-ancestors}
\end{figure}

@@

Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of only identifying one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.

While DPOR is a proved-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment. We justify the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave proving soundness to future work (and to the theoreticians).

@@

In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context-switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear. Figure~\ref{fig:found_a_bug} depicts a simple decision trace for a use-after-free bug.

>>>>

In future work (Section~\ref{sec:future-new}), the arbiter may also automatically identify extra decision points, such as conflicting shared memory accesses.

@@

At each decision point, Landslide creates a new node in the decision tree. It stores the TID of the thread that was chosen, the state of the scheduler (which threads are on the runqueue), the state of the heap and the accumulated set of shared memory accesses, and a stack trace of the thread that was running.

@@

\begin{itemize}
	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process might not yet have reaped a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well. For these cases an additional check is needed.
	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.\footnote{
		If the kernel is idling and the shell is not waiting for keyboard input, it is likely that the test case has deadlocked.}
	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
\end{itemize}

The states in the test lifecycle, and the rules for transitioning between them, are as follows.

\begin{enumerate}
	\item {\bf Bootup.} Initial state. When the kernel is idling and the shell is waiting for keyboard input, advance state.
	\item {\bf Keyboard input.} Record the current thread count (the ``start population''). Generate keystrokes to cause the desired test case to run.
		When the current thread count is greater than the start population, advance state.
	\item {\bf Test running.} Begin constructing the decision tree here (since it is pointless to backtrack to before the test begins). When the current thread count equals the start population, advance state.
	\item {\bf Test cleanup.} The test processes have exited, but some housekeeping may yet remain. When the the kernel is idling and the shell is waiting for keyboard input, advance state.
	\item {\bf Test ended.} Invoke the explorer (Section~\ref{sec:components-explore}) to decide whom to rewind, and backtrack (to step 3).
\end{enumerate}

@@

\begin{enumerate}
	\item {\bf Kernel panic bugs.} If the kernel invoked \texttt{panic}, it detected its own bug, and Landslide need do nothing but report it.
	\item {\bf Use-after-free bugs.} Whenever the kernel accesses memory in the heap (when not in the dynamic allocator itself), Landslide verifies that the address is within an allocated range. If not, Landslide proclaims the access to be illegal.\footnote{In the same way, Landslide also detects calls to \texttt{free} on blocks that were already freed or never allocated at all.}
	\item {\bf Deadlock bugs.} If Landslide finds no runnable threads on the runqueue,\footnote{Except for idle, if it exists.} or if it detects a cycle of threads blocked on each other, it declares that the kernel has deadlocked.
\suspend{enumerate}

@@

\begin{itemize}
	\item {\bf Current thread not on runqueue.} If the currently-running thread is not stored on the runqueue, we identify it as runnable anyway (except in the special case of idle, as described below).
	\item {\bf Idle thread.} Some kernels may have an explicit idle thread and store it on the runqueue, with explicit code to skip over it if other threads are runnable. Hence, if the idle thread is on the runqueue (and/or the current thread), it is only runnable if no other threads are runnable.
	\item {\bf Yielding mutexes.} In kernels whose mutexes (or other synchronisation idioms) leave threads on the runqueue when they are notionally blocked, Landslide classifies such threads as not runnable.
\end{itemize}

Using this notion of runnability, the happens-before relation is established as follows. Let $X$ be a transition and $Y$ be a subsequent transition in a branch that was just explored, and $T_Y$ be the thread associated with $Y$. We say that $X$ {\em enables} $Y$ if $T_Y$ was not runnable immediately before $X$, but was runnable immediately after. Then, a transition $A$ happens-before a transition $B$ if $T_A$ is the same thread as $T_B$, if $A$ enables $B$, or if $A$ happens-before some transition $C$ that enables $B$.

@@

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
	\subfloat[Race-free interleaving.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$C$ & & & \texttt{x=42;} \\
			$B$ & & \texttt{y=31337;} & \\
			\multirow{2}{*}{$A_1$} & \texttt{if (y==31337)} & & \\
			& \texttt{assert(x==42);} & & \\
			\cline{2-4}
		\end{tabular}
	} &
	\subfloat[Alternate interleaving explored with DPOR.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$C$ & & & \texttt{x=42;} \\
			$A_2$ & \texttt{if (y==31337)} & & \\
			$B$ & & \texttt{y=31337;} & \\
			\cline{2-4}
		\end{tabular}
	} \\
	\multicolumn{2}{c}{
	\subfloat[Buggy interleaving.]{
		\footnotesize
		\begin{tabular}{l|c|c|c|}
			\cline{2-4}
			& {\bf Thread 1} & {\bf Thread 2} & {\bf Thread 3} \\
			\cline{2-4}
			$B$ & & \texttt{y=31337;} & \\
			\multirow{2}{*}{$A_3$} & \texttt{if (y==31337)} & & \\
			& \texttt{assert(x==42);} & & \\
			$C$ & & & \texttt{x=42;} \\
			\cline{2-4}
		\end{tabular}
	}}
	\end{tabular}
	\caption{This sample code defeats a DPOR implementation that uses only the first evil ancestor if a decision point is not defined between the two lines of thread 1's code. The first interleaving is explored first; DPOR identifies transition $B$ as the first evil ancestor of $A_1$, and runs the second interleaving next. In the second interleaving, $C$ is no longer a second evil ancestor of $A_2$, so the third interleaving is never explored, and the assertion failure is missed.}
	\label{fig:evil-ancestors}
\end{figure}

@@

Figure~\ref{fig:evil-ancestors} demonstrates a counterexample to the soundness of identifying only one evil ancestor per transition given that transitions are allowed to have multiple different memory conflicts.

While DPOR is a proven-sound algorithm in environments that do have this memory access property, even with this every-ancestor change we are not certain if it is sound in our environment. We advocate the use of DPOR anyway because Landslide is already prone to false negatives because of coarse interleaving granularity, and leave investigatin soundness to future work.

@@

In its unique position of control over when the kernel gets preempted and which thread gets scheduled at each context switch, Landslide has the capability to provide the user with detailed information about a test case's execution.
When Landslide determines that a bug was found, it immediately aborts exploration of the decision tree, and prints a {\em decision trace} - a comprehensive report of the particular interleaving of thread transitions that caused the bug to appear. Figure~\ref{fig:found_a_bug} depicts a simple decision trace for a use-after-free bug.

### 2012-05-12 using.tex
more

For deadlock detection, if the kernel has mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:using-annotations}. If mutexes explicitly deschedule blocking threads, no special annotations are needed.

@@

The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation.

@@

In order to run with Landslide, the kernel must be able to boot up to the shell prompt\footnote{
In Pebbles, the bootup sequence is usually fairly simple, involving the init process spawning the shell.},
receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes. The kernel must have the following system calls implemented and working in at least a rudimentary way:
\texttt{fork}, \texttt{exec}, \texttt{vanish}, \texttt{wait}, \texttt{readline}.

@@

Here we describe the interface by which Landslide understands the execution of the guest kernel. Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.

@@

\subsubsection{Required Annotations}
\begin{itemize}
	\small
	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in the context switcher, when a new thread is switched to.
	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when the scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations.
	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)

>>>>

For deadlock detection, if the kernel has mutexes which loop around a call to \texttt{yield} (or similar), they must be annotated as described in section~\ref{sec:using-annotations}. If mutexes explicitly deschedule blocking threads, no special annotations are needed, because the other annotations do the job.

@@

The kernel should also use LMM (the \texttt{malloc} package in the 15-410 base code) for dynamic allocation, to avoid having to manually instrument a custom allocator.

@@

In order to run with Landslide, the kernel must be able to boot up to the shell prompt,\footnote{
In Pebbles, the bootup sequence is usually fairly simple, involving the init process spawning the shell.}
receive keyboard input to make the shell start a test case, and return to the shell prompt after the test case finishes. The kernel must have the following system calls implemented and working in at least a rudimentary way:
\texttt{fork}, \texttt{exec}, \texttt{vanish}, \texttt{wait}, \texttt{readline}.

@@

Here we document the interface by which Landslide understands the execution of the guest kernel. Landslide tries to be as design-agnostic as possible, after assuming that the kernel implements the Pebbles specification, but it still needs to know about the implementation of certain abstractions within the kernel.

@@

\subsubsection{Required Annotations}

Some of these annotations may need to be used in ``sensitive'' regions of scheduler code, where it may not necessarily be clear at what point during the operation the annotation should go. Landslide assumes the scheduler already protects these regions by disabling preemption, and so the exact position of the annotation does not matter as long as it is within the preemption-disabled range..

\begin{itemize}
	\small
	\item \texttt{tell\_landslide\_thread\_switch(int new\_tid)} - Call this in the context switcher, when a new thread is switched to.
	\item \texttt{tell\_landslide\_sched\_init\_done()} - Call this when the scheduler is done being initialized. Until this is called, Landslide will ignore all other annotations.
	\item \texttt{tell\_landslide\_forking()} - Call this when a new thread is being forked.
		(Hint: call it twice, once in \texttt{fork} and once in \texttt{thread\_fork}.)
